---
title: "GAE7"
author: "Katie Eufemia"
format: 
  html:
    theme: custom.scss
    embed-resources: true
    code-tools: true

csl: ecology.csl

editor: visual
bibliography: references.bib
---

```{r}
#| label: SetUp
#| message: FALSE
#| echo: FALSE

knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
  ) 

library(emmeans)
library(lmerTest)
library(redres)
library(effectsize)
library(car)
library(ggfortify)
library(patchwork)
library(broom)
library(latex2exp)       
library(kableExtra)
library(tidyverse)

options(knitr.kable.NA = '')
```

## Introduction

@Pintar.etal2018 conducted a field mesocosm experiment to explore the relative effects of predation risk and nutrient availability (a pdf of their publication is included in the GitHub repository). The predation risk manipulation was accomplished by adding either 0, 1, or 2, Golden Topminnows (*Fundulus chrysotus*) to each mesocosm. These small fish are surface-feeding predators and are known to elicit colonization and oviposition avoidance among dytiscid beetles and *Culex* mosquitoes, but not in some hydrophilid beetles. Nutrient-level treatments were created by the addition of 0, 4, or 8 g of rabbit chow (Small World Rabbit Food, Mannapro, St. Louis, MO, 40% protein).

## Methods

Please see @Pintar.etal2018 for greater detail on experimental design and data collection. In brief, the experiment was set up as a 3 X 3 factorial treatment design (9 treatment combinations). The experiment was replicated in 3 separate locations at 3 different times for a total of 9 replicates. This should have resulted in 81 observations total, however some fish died during the experiment rendering those observations not being included in the data resulting in only 69 observations.

```{r }


df1 <- read_csv("Pintar_etal_2018.csv",
                col_select = c(1:9),
                show_col_types = FALSE
                )


# Block = 1 to 9 for the 9 spatiotemporal blocks (3 locations X 3 times)
# Pool = 1 to 9 for the plastic wading pools used for the experiment
# Location = 1 to 3 for the 3 fields where the pools were deployed
# Time = a to c for the 3 time periods used for each 2-week run of the exp.
# Nutrient = 1 to 3 for the 3 nutrient levels
# Fish = 1 to 3 for the 3 fish densities
# Culex_Egg_Rafts = number of egg rafts collected over the 2 weeks
# Hydrophilidae = number of hydrophilid beetles (Water Scavenger Beetles)
# Dytiscidae = number of dytiscid beetles (Predaceous Diving Beetles)

# Fish and Nutrient are provided as integers, but I wish to use them as ordered
# factors so that, by default, polynomial contrasts will be used to compute the
# ANOVAs.

df1$Block <- factor(df1$Block)

df1$Pool <- factor(df1$Pool)

df1$Location <- factor(df1$Location)

df1$Time <- factor(df1$Time)

df1$Nutrient <- factor(df1$Nutrient)

df1$Fish <- factor(df1$Fish)


```

```{r}
#| label: tbl-dataImport
#| tbl-cap: "First six observations of data provided by @Pintar.etal2018."

colnames(df1) <- tolower(colnames(df1))


kbl(
  head(df1),
  col.names = c(
    "Block",
    "Pool",
    "Location",
    "Time",
    "Nutrient",
    "Fish",
    "Culex Egg Rafts",
    "Hydrophilidae",
    "Dytiscidae"
  )
) |>
  kable_classic(
    full = FALSE
    )
```

### Statistical Methods

The response variable was total number of egg rafts based on the different treatments and varying nutrient availability. The 9 treatments with separate locations and times were considered blocks and the different spatio-temporal blocks were assumed independent. Blocks were also assumed to have normal independent deviance with a mean of zero. The data were analyzed with the blocks as fixed:

$$
y_{ij} = \mu + \alpha_i + \beta_j + \epsilon_{ij}$$

## Results

```{r}
#| label: tbl-Culex
#| tbl-cap: "Effects of predation and nutrients on Culex."
#| fig-show: hide
#| results: hide

CulexFactorial <- lmer(culex_egg_rafts ~ fish*nutrient + (1 | block), data = df1)

CulexAnova <- (anova(CulexFactorial))

kbl(
  CulexAnova,
  col.names = c(
    "SS",
    "MS",
    "df1",
    "df2",
    "F",
    "P"
    ),
  digits = c(3,3,0,0,3,4)
  ) |>
  kable_classic()

HydroFactorial <- lmer(hydrophilidae ~ fish*nutrient + (1 | block), data = df1)

HydroAnova <- (anova(HydroFactorial))

kbl(
  HydroAnova,
  col.names = c(
    "SS",
    "MS",
    "df1",
    "df2",
    "F",
    "P"
    ),
  digits = c(3,3,0,0,3,4)
  ) |>
  kable_classic()

DytisFactorial <- lmer(dytiscidae ~ fish*nutrient + (1 | block), data = df1)

DytisAnova <- (anova(DytisFactorial))

kbl(
  DytisAnova,
  col.names = c(
    "SS",
    "MS",
    "df1",
    "df2",
    "F",
    "P"
    ),
  digits = c(3,3,0,0,3,4)
  ) |>
  kable_classic()

```

```{r}
#| label: tbl-fixedANOVA
#| tbl-cap: "ANOVA summary for the data provided by @Pintar.etal2018."
#| fig-show: hide

kbl(
  anova(CulexFactorial),
  digits = c(3,0,3,3)
  ) |>
  kable_classic()
```

```{r}
#| label: tbl-fixedANOVA2
#| tbl-cap: "ANOVA summary for the data provided by @Pintar.etal2018."
#| fig-show: hide

kbl(
  anova(HydroFactorial),
  digits = c(3,0,3,3)
  ) |>
  kable_classic()
```

```{r}
#| label: tbl-fixedANOVA3
#| tbl-cap: "ANOVA summary for the data provided by @Pintar.etal2018."
#| fig-show: hide

kbl(
  anova(DytisFactorial),
  digits = c(3,0,3,3)
  ) |>
  kable_classic()
```

Because blocks were chosen at random, they should be treated as a random factor.

```{r}
#| label: tbl-mixedANOVA
#| tbl-cap: "Linear mixed-effect model results for the data provided by @Pintar.etal2018."
#| tbl-subcap:
#|     - "Fixed effect."
#|     - "Random effect."
#| layout-nrows: 2


CulexMixed <- anova(CulexFactorial)

kbl(
  CulexMixed,
  col.names = c(
    "SS",
    "MS",
    "df1",
    "df2",
    "F",
    "P"
    ),
  digits = c(3,3,0,0,3,4)
  ) |>
  kable_classic()

kbl(
  rand(CulexFactorial),
  col.names = c(
    "Parameters",
    "log-likelihood",
    "AIC",
    "LRT",
    "df",
    "P"
    ),
  digits = c(0,3,3,3,0,4)
  )
```

```{r}
#| label: tbl-mixedANOVA2
#| tbl-cap: "Linear mixed-effect model results for the data provided by @Pintar.etal2018."
#| tbl-subcap:
#|     - "Fixed effect."
#|     - "Random effect."
#| layout-nrows: 2


HydroMixed <- anova(HydroFactorial)

kbl(
  HydroMixed,
  col.names = c(
    "SS",
    "MS",
    "df1",
    "df2",
    "F",
    "P"
    ),
  digits = c(3,3,0,0,3,4)
  ) |>
  kable_classic()

kbl(
  rand(HydroFactorial),
  col.names = c(
    "Parameters",
    "log-likelihood",
    "AIC",
    "LRT",
    "df",
    "P"
    ),
  digits = c(0,3,3,3,0,4)
  )
```

```{r}
#| label: tbl-mixedANOVA3
#| tbl-cap: "Linear mixed-effect model results for the data provided by @Pintar.etal2018."
#| tbl-subcap:
#|     - "Fixed effect."
#|     - "Random effect."
#| layout-nrows: 2


DytisMixed <- anova(DytisFactorial)

kbl(
  DytisMixed,
  col.names = c(
    "SS",
    "MS",
    "df1",
    "df2",
    "F",
    "P"
    ),
  digits = c(3,3,0,0,3,4)
  ) |>
  kable_classic()

kbl(
  rand(DytisFactorial),
  col.names = c(
    "Parameters",
    "log-likelihood",
    "AIC",
    "LRT",
    "df",
    "P"
    ),
  digits = c(0,3,3,3,0,4)
  )
```

## Discussion

Because of the insignificant difference in Akaike's Information Criterion, there is little difference between the models (@tbl-mixedANOVA, @tbl-mixedANOVA2, @tbl-mixedANOVA3) . The analyses also show that Culex mosquitoes have a significant interaction with nutrients and predator abundance while dytiscid and hydrophilid beetles do not (@tbl-fixedANOVA, @tbl-fixedANOVA2, @tbl-fixedANOVA3).

## References
